---
title: "Aim2_AMR_Analysis"
author: "Emily Wissel"
date: "1/31/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gee) # for significance test at the end
```

## Catalogue of AMR Genes Detected during Pregnancy 

Read in the meta data
```{r meta data}
#read in medical data
med_dat <- read_csv("../medical_data/MPTB_Sociodemographic Outcome Health Behaviors update4, 16S, WGS indicator.csv")
med_dat <- med_dat %>% filter(Sequencing_WGS == "1"  )

# cofounder variables: age, income, parity, TobaccoUse_MRm AlcoholUse_MR, MarijuanaUse_MR, 
# covariate variables: birthoutcome, labor, Preg_Chlam, Preg_UTI

# merge med_dat and small
med_dat$subjectID <- med_dat$subjectid
small <- dat_otu %>% select(sample, subjectID, Plate, timepoint, bodysite) # from aim 1 r script
small <- small [order(small$sample),]
med_data <- merge(small, med_dat, all = TRUE, by.x = "subjectID", by.y = "subjectid" )
#med_data # 937 rows
med_data[order(med_data$sample),] ## motu and med_dat should have same order of samples in rows
#write.csv(med_dat, "med_data_for_mj.csv")

med_data$sample<-gsub("_profile","",as.character(med_data$sample))

med_dat_vag <- med_data %>% filter(bodysite == "vaginal")
med_dat_rectal <- med_data %>% filter(bodysite == "vaginal")

table(med_data$Preg_BV)
table(med_data$Preg_UTI)
table(med_data$Preg_Chlam)
#med_dat_vag #%>% unique(subjectID)

length(unique(med_dat$subjectID) )

head(med_dat_vag)
```



Firest we read in the data
```{r read in data}
dat <- read_csv("compiled_amrfinderplus_output.csv",  col_names = FALSE)
colnames(dat) <- c("sample", "gene_length", "gene_name", "drug_class")
dat$gene_name_lower <- tolower(dat$gene_name)

# replace weird na with proper na
#dat[is.na(dat)] <- "NA"
length(unique(dat$sample) ) # 814 
table(dat$drug_class)


df <- left_join(dat, med_data)
## remove pp samples 
df <- df %>%filter(timepoint != "NA")

## inspect AMR genes in control samples
#df %>% filter(bodysite == "control")

# get median amr genes per sample per bodysite
df %>%
  filter(bodysite != "control") %>%
  group_by(subjectID, bodysite) %>%
  count(drug_class) %>%
  ungroup %>%
  group_by(bodysite) %>%
  summarize(Mean=mean(n), Max=max(n), Min=min(n), Median=median(n), Std=sd(n))

vag_df <- dat %>% 
  mutate(bodysite = ifelse(str_detect(sample, 'Vag'), "vaginal", 
                           ifelse(str_detect(sample, "Rec"), "rectal", 
                                  "control"))) %>%
  mutate(timepoint = ifelse(str_detect(sample, "-1-"), "1",
                            ifelse(str_detect(sample, "-2-"), "2", "other"))) %>%
  filter(bodysite == "vaginal") %>% 
  group_by(drug_class) %>%
  filter(n()>10) %>% # filter out drug classes observed in fewer than 10 samples
  ungroup()
vag_df # 1,831 rows
vag_df <- left_join(vag_df, med_data)

vag_df %>%
  filter(bodysite != "control") %>%
  group_by(subjectID, bodysite) %>%
  count(drug_class) %>%
  ungroup %>%
  group_by(bodysite) %>%
  summarize(Mean=mean(n), Max=max(n), Min=min(n), Median=median(n), Std=sd(n))


```

nOW MAKE A VISUAL
```{r}
tidydat <- df %>%
 mutate(infxn_id = ifelse(Preg_Chlam == "1" | Preg_BV == "1" | Preg_UTI == "1", "vag_infxn",
                           ifelse(Antibiotic_oral == "1" | OralAbxBetween1.2 == "1" | ParenAbxBetween1.2 == "1" |
                                    Antibiotic_parenteral == "1" | Antibiotic_vaginal == "1" | Antifungal_oral == "1" |
                                    Antifungal_vaginal == "1", 
                                  "other_abx_infxn", "no_infxn") ))  %>%
  group_by(subjectID, bodysite, infxn_id, timepoint) %>%
  count(drug_class)

check <- med_data %>%
  mutate(infxn_id = ifelse(Preg_Chlam == "1" | Preg_BV == "1" | Preg_UTI == "1", "vag_infxn",
                           ifelse(Antibiotic_oral == "1" | OralAbxBetween1.2 == "1" | ParenAbxBetween1.2 == "1" |
                                    Antibiotic_parenteral == "1" | Antibiotic_vaginal == "1" | Antifungal_oral == "1" |
                                    Antifungal_vaginal == "1", 
                                  "other_abx_infxn", "no_infxn") )) 

check %>%
  select(sample, subjectID, infxn_id, Preg_Chlam, Preg_UTI, Preg_BV, Antibiotic_oral, OralAbxBetween1.2, ParenAbxBetween1.2, Antibiotic_parenteral, Antibiotic_vaginal, Antifungal_oral, Antifungal_vaginal) %>%
  filter(infxn_id == "NA")
## this is working the way it is suppose to, yay! 



tidydat %>%
  ggplot(aes(x = drug_class, y = n, fill = infxn_id, color = infxn_id) ) + 
  geom_boxplot() + 
  geom_jitter(height = 0.1) +
  facet_wrap(.~timepoint) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) 
  
tidydat %>%
  filter(infxn_id != "NA") %>%
  ggplot(aes(x = drug_class, y = n,  color = infxn_id) ) + 
  geom_boxplot() + 
  geom_jitter(height = 0.1, width = 0.2) +
  facet_wrap(.~infxn_id, ncol = 1) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_color_manual(values = c( "lavenderblush3", "thistle", "lightsteelblue2")) +
  theme_light() +
  labs(title = " AMR genes detected per sample across drug classes")
```


now we set it up for a chi squared test. we should use the Fisher’s exact test instead of the Chi-square test because there is at least one cell below 5.

```{r chi squared}
predat <- tidydat %>%
  ungroup() %>%
  filter(infxn_id != "NA") %>%
  group_by(bodysite, infxn_id, drug_class, timepoint) %>%
  summarize(amr_count = sum(n))
  
rectal_agg_dat <-predat %>%
  #pivot_wider(names_from = "drug_class", values_from = 'amr_count', values_fill = 0) %>%
  filter(bodysite == "vaginal")%>%
  ungroup () %>%
  select(-bodysite)
# make a matrix
#rownames(rectal_agg_dat) <- rectal_agg_dat$infxn_id
rec_agg_mat <- as.matrix(rectal_agg_dat)
#rectal_agg_dat

vag_agg_dat <- predat %>%
  #pivot_wider(names_from = "drug_class", values_from = 'amr_count', values_fill = 0) %>%
  filter(bodysite == "vaginal")%>%
  ungroup () %>%
  select(-bodysite)

#View(rectal_agg_dat)

## make figure to show contingency table


rectal_agg_dat %>%
 ggplot(aes(x = drug_class, y = infxn_id)) +
  geom_point(aes(size = amr_count), shape = 21, colour = "black", fill = "cornsilk") +
  scale_size_area(max_size = 20, guide = FALSE) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  geom_text(aes(
    y = infxn_id, label = amr_count),
    vjust = 3.5,
    colour = "grey60",
    size = 4) +
  labs(title = "Rectal AMR Genes")
vag_agg_dat %>%
 ggplot(aes(x = drug_class, y = infxn_id)) +
  geom_point(aes(size = amr_count), shape = 21, colour = "black", fill = "cornsilk") +
  scale_size_area(max_size = 20, guide = FALSE) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  geom_text(aes(
    y = infxn_id, label = amr_count),
    vjust = 3.5,
    colour = "grey60",
    size = 4) +
  labs(title = "Vaginal AMR Genes")

#compile
predat %>%
 ggplot(aes(x = drug_class, y = infxn_id, fill = infxn_id)) +
  geom_point(aes(size = amr_count), shape = 21, colour = "black") +
  scale_size_area(max_size = 20, guide = FALSE) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  geom_text(aes( y = infxn_id, label = amr_count),
    vjust = 3.5, colour = "grey60",size = 4) +
  labs(title = "AMR Genes") +
  facet_wrap(.~bodysite, ncol = 1) +
  scale_fill_manual(values = c( "lavender", "thistle", "lightsteelblue2"))

predat
```

lets do a chi squared for each bodysite at each timepoint. The chi-squared test helps to determine whether there is a notable difference between the normal frequencies and the observed frequencies in one or more classes or categories. It gives the probability of independent variables.

```{r chiu squared rectal}
rec_df <- dat %>% 
  mutate(bodysite = ifelse(str_detect(sample, 'Vag'), "vaginal", 
                           ifelse(str_detect(sample, "Rec"), "rectal", 
                                  "control"))) %>%
  mutate(timepoint = ifelse(str_detect(sample, "-1-"), "1",
                            ifelse(str_detect(sample, "-2-"), "2", "other"))) %>%
  filter(bodysite == "rectal") %>% 
  group_by(drug_class) %>%
  filter(n()>10) %>% # filter out drug classes observed in fewer than 10 samples
  ungroup()
rec_df # 9,417 rows
rec_df <- left_join(rec_df, med_data)

rectal_agg_dat_t1 <- rec_df %>%
 mutate(infxn_id = ifelse(Preg_Chlam == "1" | Preg_BV == "1" | Preg_UTI == "1", "vag_infxn",
                           ifelse(Antibiotic_oral == "1" | OralAbxBetween1.2 == "1" | ParenAbxBetween1.2 == "1" |
                                    Antibiotic_parenteral == "1" | Antibiotic_vaginal == "1" | Antifungal_oral == "1" |
                                    Antifungal_vaginal == "1", 
                                  "other_abx_infxn", "no_infxn") ))  %>%
  ungroup() %>%
  filter(infxn_id != "NA") %>%
  group_by(bodysite, infxn_id, drug_class, timepoint) %>%
  filter(bodysite == "rectal") %>% 
  filter(timepoint == "1") %>%
  ungroup() %>% group_by(drug_class) %>%
  filter(n()>10) %>% # filter out drug classes observed in fewer than 10 samples
  ungroup()


rectal_agg_dat_t2 <- rec_df %>%
 mutate(infxn_id = ifelse(Preg_Chlam == "1" | Preg_BV == "1" | Preg_UTI == "1", "vag_infxn",
                           ifelse(Antibiotic_oral == "1" | OralAbxBetween1.2 == "1" | ParenAbxBetween1.2 == "1" |
                                    Antibiotic_parenteral == "1" | Antibiotic_vaginal == "1" | Antifungal_oral == "1" |
                                    Antifungal_vaginal == "1", 
                                  "other_abx_infxn", "no_infxn") ))  %>%
  ungroup() %>%
  filter(infxn_id != "NA") %>%
  group_by(bodysite, infxn_id, drug_class, timepoint) %>%
  filter(bodysite == "rectal") %>% 
  filter(timepoint == "2") %>%
  ungroup()%>% group_by(drug_class) %>%
  filter(n()>10) %>% # filter out drug classes observed in fewer than 10 samples
  ungroup()



############ run those tests bb
t1_rec <- chisq.test(table(rectal_agg_dat_t1$infxn_id, rectal_agg_dat_t1$drug_class))
t1_rec$observed # BETA-LACTAM MACROLIDE TETRACYCLINE VIRULENCE
t1_rec$p.value
t1_rec$residuals
t1_rec$stdres

t2_rec <- chisq.test(table(rectal_agg_dat_t2$infxn_id, rectal_agg_dat_t2$drug_class))
t2_rec$observed # BETA-LACTAM MACROLIDE TETRACYCLINE VIRULENCE
t2_rec$p.value # no sig difference 
t2_rec$stdres

intm <-df %>% mutate(infxn_id = ifelse(Preg_Chlam == "1" | Preg_BV == "1" | Preg_UTI == "1", "vag_infxn",
                           ifelse(Antibiotic_oral == "1" | OralAbxBetween1.2 == "1" | ParenAbxBetween1.2 == "1" |
                                    Antibiotic_parenteral == "1" | Antibiotic_vaginal == "1" | Antifungal_oral == "1" |
                                    Antifungal_vaginal == "1", 
                                  "other_abx_infxn", "no_infxn") ))  %>%
  select(subjectID, infxn_id) %>%
  distinct()

table(intm$infxn_id)
#unique(rectal_agg_dat_t2$drug_class)


```
interp guide: https://scholarworks.umass.edu/cgi/viewcontent.cgi?article=1269&context=pare
look at standardized residuals greater than |2|. 

now the same for vaginal data
```{r chi squared vag}
vag_agg_dat_t1 <- vag_df %>%
 mutate(infxn_id = ifelse(Preg_Chlam == "1" | Preg_BV == "1" | Preg_UTI == "1", "vag_infxn",
                           ifelse(Antibiotic_oral == "1" | OralAbxBetween1.2 == "1" | ParenAbxBetween1.2 == "1" |
                                    Antibiotic_parenteral == "1" | Antibiotic_vaginal == "1" | Antifungal_oral == "1" |
                                    Antifungal_vaginal == "1", 
                                  "other_abx_infxn", "no_infxn") ))  %>%
  ungroup() %>%
  filter(infxn_id != "NA") %>%
  group_by(bodysite, infxn_id, drug_class, timepoint) %>%
  filter(bodysite == "vaginal") %>% 
  filter(timepoint == "1") %>%
  ungroup() %>% group_by(drug_class) %>%
  filter(n()>10) %>% # filter out drug classes observed in fewer than 10 samples
  ungroup()


vag_agg_dat_t2 <- vag_df%>%
 mutate(infxn_id = ifelse(Preg_Chlam == "1" | Preg_BV == "1" | Preg_UTI == "1", "vag_infxn",
                           ifelse(Antibiotic_oral == "1" | OralAbxBetween1.2 == "1" | ParenAbxBetween1.2 == "1" |
                                    Antibiotic_parenteral == "1" | Antibiotic_vaginal == "1" | Antifungal_oral == "1" |
                                    Antifungal_vaginal == "1", 
                                  "other_abx_infxn", "no_infxn") ))  %>%
  ungroup() %>%
  filter(infxn_id != "NA") %>%
  group_by(bodysite, infxn_id, drug_class, timepoint) %>%
  filter(bodysite == "vaginal") %>% 
  filter(timepoint == "2") %>%
  ungroup()%>% group_by(drug_class) %>%
  filter(n()>10) %>% # filter out drug classes observed in fewer than 10 samples
  ungroup()  



############ run those tests bb
t1_vag <- chisq.test(table(vag_agg_dat_t1$infxn_id, vag_agg_dat_t1$drug_class))
t1_vag$observed # BETA-LACTAM MACROLIDE TETRACYCLINE VIRULENCE
t1_vag$p.value # no sig difference
t1_vag

t2_vag <- chisq.test(table(vag_agg_dat_t2$infxn_id, vag_agg_dat_t2$drug_class))
t2_vag$observed # BETA-LACTAM MACROLIDE TETRACYCLINE VIRULENCE
t2_vag$p.value #  sig difference 
t2_vag$stdres
t2_vag$residuals


```

### Generalized Estimating Equiation (GEE)

(source)[https://data.library.virginia.edu/getting-started-with-generalized-estimating-equations/]

This is a good nonparametric approach for "simple"(er) modelling with longitudinal and/or clustered data. Dr. Melinda Higgins recommended I take this approach when I asked about longitudinal data and chi squared tests.


```{r gee}
gee_dat <- tidydat %>%
  ungroup() %>%
  filter(infxn_id != "NA") %>%
  #select(sample, subjectID, bodysite, timepoint, infxn_id, drug_class) %>%
  na.omit() %>% group_by(drug_class) %>%
  filter(n()>10) %>% # filter out drug classes observed in fewer than 10 samples
  ungroup()

gee_dat$drug_class <- as.factor(gee_dat$drug_class)
gee_dat$infxn_id <- as.factor(gee_dat$infxn_id)
gee_dat$timepoint <- as.factor(gee_dat$timepoint)
gee_dat$subjectID <- as.factor(gee_dat$subjectID)
#######
gee_rec <- gee_dat %>% filter(bodysite == "rectal") %>% arrange(subjectID) ## gee assumes clustered data is in continous rows
gee_vag = gee_dat %>%filter(bodysite == "vaginal")%>% arrange(subjectID)
########### set up complete
gtest_rec <- df %>%
 mutate(infxn_id = ifelse(Preg_Chlam == "1" | Preg_BV == "1" | Preg_UTI == "1", "vag_infxn",
                           ifelse(Antibiotic_oral == "1" | OralAbxBetween1.2 == "1" | ParenAbxBetween1.2 == "1" |
                                    Antibiotic_parenteral == "1" | Antibiotic_vaginal == "1" | Antifungal_oral == "1" |
                                    Antifungal_vaginal == "1", 
                                  "other_abx_infxn", "no_infxn") ))  %>%
  group_by(subjectID, bodysite, infxn_id, timepoint) %>%
  filter(infxn_id != "NA") %>%
  select(sample, subjectID, bodysite, timepoint, infxn_id, drug_class) %>%
  na.omit() %>%
  filter(bodysite=="rectal") %>%
  ungroup()
table(gtest_rec$drug_class, gtest_rec$infxn_id)

###############
form = n ~ drug_class + infxn_id + timepoint

gout_rec <- gee(formula = form,
    id = subjectID,
    data = gee_rec,
    family = "poisson"
    )

#predat ## all samples pooled
summary(gout_rec)
gout_rec
```

heko wiht interp? https://stats.stackexchange.com/questions/215980/how-to-interpret-geeglm-results 

let's just replicate in the vaginal data and figure out what it means later because i am severely sleep deprived 

```{r all bodysites gee}
## filter so that only drug classes with at least 10 observed AMR genes across the whole dataset are included 

###############
form_whole = n ~ drug_class + infxn_id + timepoint + bodysite

gout_whole <- gee(formula = form_whole,
    id = subjectID,
    data = gee_dat,
    family = "poisson"
    )

#predat ## all samples pooled
summary(gout_whole)
gout_whole
###### get p values
#coef(summary(gout_whole))
2 * pnorm(abs(coef(summary(gout_whole))[,5]), lower.tail = FALSE)
```

## compare diversity across infxn_id groups

```{r alpha diversity}
head(data_alphadiv)


wide_dat <- tidydat %>%
  pivot_wider(names_from = drug_class, values_from = n,  values_fill = 0)
wide_dat ## 727
alph_dat <- left_join(wide_dat, data_alphadiv)
alph_dat <- alph_dat %>%
  filter(timepoint != "control" ) %>%
  filter(infxn_id != "NA")
alph_dat$timepoint <- as.numeric(alph_dat$timepoint)

alph_dat %>%
  ggplot(aes(x = timepoint, y = data_shannon, fill = infxn_id)) +
  geom_boxplot(outlier.shape = NA) + 
  geom_point(position=position_jitterdodge(jitter.width =  0.2), alpha = 0.5) + 
  facet_wrap(.~bodysite) +
  theme_light() +
  labs(title = "Shannon Diversity",y = "Shannon Diversity", x = "Time Point") +
  scale_fill_manual(values = c( "lavenderblush3", "thistle", "lightsteelblue2")) 


summary(aov(data_shannon | (timepoint)~ infxn_id, data = alph_dat))
#### rectal only
alph_dat_rec <- alph_dat %>% filter(bodysite == "rectal")
describe(alph_dat_rec$data_shannon)
summary(aov(data_shannon | (timepoint)~ infxn_id, data = alph_dat_rec))
#### rectal only
alph_dat_vag <- alph_dat %>% filter(bodysite == "vaginal")
summary(aov(data_shannon | (timepoint)~ infxn_id, data = alph_dat_vag))
describe(alph_dat_vag$data_shannon)
```

